---
title: "PopHIVE: Population Health Information and Visualization Exchange"
title-block-banner: "#0c233f"
format:
  html:
    embed-resources: true
    page-layout: full
    toc: true
    toc-expand: true
    smooth-scroll: true
    theme: cosmo
css: styles.css
server: shiny
---

```{r setup, include=FALSE}
#| context: setup
#| include: false

library(tidyverse)
library(plotly)
library(shiny)
library(tidyverse)
library(scales)
library(janitor)
library(MMWRweek)
library(arrow)
#library(rnaturalearth)
library(parquetize)
library(viridisLite)
library(tigris)
library(usmap)
library(cowplot)
library(viridis)
library(sf)
library(ggrepel)
library(readxl)
library(ggalluvial)
library(waffle)
#library(cdcfluview)
#install.packages("remotes")
#remotes::install_github("hrbrmstr/cdcfluview")
#hhs_regions$state <- state.abb[match(hhs_regions$state_or_territory,state.name)]
#state.hhs.codes <- aggregate(hhs_regions$state, #list(hhs_regions$region), paste, collapse=",")
#state.hhs.codes$x <- gsub('NA,','',state.hhs.codes$x)
#state.hhs.codes$x <- gsub(',NA','',state.hhs.codes$x)
#saveRDS(state.hhs.codes, './hhs_regions.rds')

yale_colors <- c(
  "#00356B",
  "#286DC0",
  "#63AAFF",
  "#C4DDFC",
  "#DDDDDD"
)

yale_gradient <- c("#DDDDDD",
                   "#C4DDFC",
                   "#63AAFF",
                   "#286DC0",
                   "#00356B") # Adjust order for your desired gradient



#datasets for use in the global environment


"%!in%" <- function(x,y)!("%in%"(x,y))

# states <- ne_states(country = "United States of America", returnclass = "sf")
# state_centers <- centers <- st_centroid(states) |> 
#   st_coordinates() |> 
#   as.data.frame() |>
#   cbind(State = states$name)
```

```{r rsv_data}
#| context: data
#| include: false
#| cache: true
#| cache.extra: !expr file.info("data.csv")$mtime


#csv_to_parquet('https://raw.githubusercontent.com/ysph-dsde/data-gov/refs/heads/main/RSV%20Infections%20Data/Raw%20Download/NREVSS%20by%20HHS%20Region_Respiratory%20Syncytial%20Virus%20(RSV)%20Surveillance%20Archive_Downloaded%2001.23.2025.csv',path_to_parquet ='./Data/rsv_testing.parquet')

rsv1_tests <- open_dataset('./Data/rsv_testing.parquet') %>%
  collect() %>%
  as.data.frame()


key <- readRDS('./Data/hhs_regions.rds')

rsv_ts <- rsv1_tests %>%
  mutate(date= as.Date(substr(mmwrweek_end,1,10 ), '%m/%d/%Y'),
         postdate= as.Date(substr(posted,1,10 ), '%m/%d/%Y')) %>%
  filter(postdate==max(postdate)) %>%
  ungroup() %>%
  filter(level != 'National') %>%
  group_by(level ) %>%
  left_join(key, by=c('level'='Group.1')) %>%
  mutate(scaled_cases = pcr_detections/max(pcr_detections),
         hhs_abbr = x  ) %>%
  ungroup()

dates2_rsv_ts <- MMWRweek(as.Date(rsv_ts$date))

max.wk.yr <- max(dates2_rsv_ts$MMWRweek[dates2_rsv_ts$MMWRyear==max(dates2_rsv_ts$MMWRyear)])

rsv_ts <- cbind.data.frame(rsv_ts,dates2_rsv_ts[,c('MMWRyear', 'MMWRweek')]) %>%
  mutate( epiyr = MMWRyear, 
          epiyr = if_else(MMWRweek<=26,MMWRyear - 1 ,MMWRyear),
          epiwk  = if_else( MMWRweek<=26, MMWRweek+52, MMWRweek  ),
          epiwk=epiwk-26
  )

#Detailed data, did a static download; updated regularly via API rdmq-nq56
##convert from CSV to parquet, which is much more compact!
# csv_to_parquet(path_to_file = './Data/CopyOfNSSP_Emergency_Department_Visit_Trajectories_by_State_and_Sub_State_Regions-_COVID-19__Flu__RSV__Combined___20250130.csv', path_to_parquet ='./Data/NSSP_detailed.parquet') 
dates <- seq.Date(from=as.Date('2022-10-01'), to=as.Date('2025-01-30'),by='week')

#TODO fix county mapping in CT https://stackoverflow.com/questions/78599492/plotting-old-connecticut-county-regions-in-usmap
#for(i in 1:length(dates)){
#}

# 
# d1_all <- open_dataset('./Data/NSSP_detailed.parquet') %>%
#   filter(county!='All') %>%collect()


# Waste water Latest: https://www.cdc.gov/wcms/vizdata/NCEZID_DIDRI/RSVStateLevelDownloadCSV.csv

#hospitalizations
#
# csv_to_parquet('https://raw.githubusercontent.com/ysph-dsde/data-gov/refs/heads/main/RSV%20Infections%20Data/Harmonized%20RSV-NET_01.29.2025.csv',path_to_parquet ='./Data/rsvnet_hosp.parquet')


# csv_to_parquet('./Data/NSSP_demographics.csv',path_to_parquet ='./Data/NSSP_demographics.parquet')

h1 <- open_dataset('./Data/rsvnet_hosp.parquet') %>%
  filter(`Region Type`=='State' & `Diagnostic Test Type`=='All' & Characteristic=='Not Stratified' & Region != 'All Sites') %>%
  rename(state=Region, hosp_rate=`Crude Rate`, date=`Week Observed`) %>%
  dplyr::select(state, date, hosp_rate) %>%
  collect() %>%
  as.data.frame()

h1.age <- open_dataset('./Data/rsvnet_hosp.parquet') %>%
  filter(`Region Type`=='State' & `Diagnostic Test Type`=='All' & Characteristic=='Age'  & Region != 'All Sites') %>%
  rename(state=Region, hosp_rate=`Crude Rate`, date=`Week Observed`) %>%
  dplyr::select(state, date, hosp_rate, Level) %>%
  collect() %>%
  ungroup() %>%
  filter( date >=as.Date('2023-07-01')) %>%
  group_by(state, Level) %>%
  mutate(scale_age=hosp_rate/max(hosp_rate, na.rm=T )) %>%
  as.data.frame()

#Wastewaster by state and date
w1 <- read.csv('./Data/RSVStateLevelDownloadCSV_ww.csv') %>%
  mutate(date=as.Date(Week_Ending_Date, '%m/%d/%Y')) %>%
  filter(Data_Collection_Period=='All Results') %>%
  rename(state=State.Territory, rsv_ww=State.Territory_WVAL) %>%
  arrange(state, date) %>%
  dplyr::select(state, date, rsv_ww)


#ED visits by state overall
d1 <- open_dataset('./Data/NSSP_detailed.parquet') %>%
  filter(county=='All'  ) %>%
  rename(state=geography, date='week_end') %>%
  dplyr::select(state, date, percent_visits_rsv) %>%
  collect() %>%
  as.data.frame()

#combine hospital, wastewater, ED visits
dwh <- d1 %>%
  full_join(w1, by=c('state','date')) %>%
  full_join(h1, by=c('state', 'date'))

dates2 <- MMWRweek(as.Date(dwh$date))

max.wk.yr <- max(dates2$MMWRweek[dates2$MMWRyear==max(dates2$MMWRyear)])

dwh <- cbind.data.frame(dwh,dates2[,c('MMWRyear', 'MMWRweek')]) %>%
  mutate( epiyr = MMWRyear, 
          epiyr = if_else(MMWRweek<=26,MMWRyear - 1 ,MMWRyear),
          epiwk  = if_else( MMWRweek<=26, MMWRweek+52, MMWRweek  ),
          epiwk=epiwk-26
  )


#ED visits by demographics
# d1 <- open_dataset('./Data/NSSP_demographics.parquet') %>%
#   collect()
#   filter(county=='All'  ) %>%
#   rename(state=geography, date='week_end') %>%
#   dplyr::select(state, date, percent_visits_rsv) %>%
#   collect() %>%
#   as.data.frame()
```

```{r pneumococcus_data}
#| context: data
#| include: false
#| cache: true
#| cache.extra: !expr file.info("data.csv")$mtime


###Pneumococcal disease
#csv_to_parquet('https://data.cdc.gov/resource/qvzb-qs6p.csv',path_to_parquet ='./Data/ipd_cdc1998.parquet')

ipd1 <- readRDS('./Data/pneumococcus/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st),
          agec1 = if_else(agec %in% c("Age <2","Age 2-4") ,1,2 ),
          agec=gsub('Age ', '', agec),
          agec2 = if_else( agec %in% c('<2','2-4'), '<5',
                           if_else( agec %in% c('5-17','18-49'), '5-49',
                                    if_else( agec %in% c('50-64','65+'), '50+',NA))),
          agec2 = factor(agec2, levels=c('<5','5-49','50+'), labels=c('<5 years', '5-49 years', '50+ years') )
  ) %>%
  group_by(st,agec2, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup()

all.sts <- unique(ipd1$st)


# pneumococcal serotype by state
b2019 <- read.csv('./Data/pneumococcus/jiac058_suppl_supplementary_table_s2.csv') %>%
  group_by(State, sero) %>%
  summarize(N_cases=n()) %>%
  mutate(sero=as.factor(sero)) %>%
  ungroup() %>%
  group_by(State, sero) %>%
  mutate(mean_cases=max(N_cases,na.rm=T)
  ) %>%
  group_by(State) %>%
  mutate(         pct = N_cases/sum(N_cases, na.rm=T)*100) %>%
  ungroup() %>%
  tidyr::complete(sero,State , fill=list(pct=0))  #fills 0
```

```{r vax_data}
#| context: data
#| include: false
#| cache: true
#| cache.extra: !expr file.info("data.csv")$mtime

#pediatric vaccine yptake
#https://data.cdc.gov/Child-Vaccinations/Vaccination-Coverage-among-Young-Children-0-35-Mon/fhky-rtsk/about_data

# csv_to_parquet('./Data/vax/Vaccination_Coverage_among_Young_Children__0-35_Months__20250204.csv',path_to_parquet ='./Data/vax/peds_vax.parquet')
#this breaks out the age data; also has info on income, race, urbanicity, etc
vax_age <- open_dataset('./Data/vax/peds_vax.parquet') %>%
  rename(birth_year = `Birth Year/Birth Cohort`, dim1=`Dimension Type`, age=Dimension,vax_uptake=`Estimate (%)`, samp_size_vax=`Sample Size`) %>%
  collect() %>%
  filter(birth_year==2021 & dim1=='Age') %>%
  dplyr::select(Vaccine,Geography, Dose, dim1, vax_uptake,samp_size_vax, age) %>%
  filter(age=='35 Months' & Geography %in% state.name &
           (Vaccine %in% c('≥1 Dose MMR','≥1 Dose Varicella' ) | 
              (Vaccine=='DTaP' & Dose =='≥4 Doses') | 
              (Vaccine=='Hep A' & Dose =='≥2 Doses') | 
              (Vaccine=='Hep B' & Dose =='≥3 Doses') | 
              (Vaccine=='Hib' & Dose =='Full Series') | 
              (Vaccine=='PCV' & Dose =='≥4 Doses') 
           ) )

vax_urban <- open_dataset('./Data/vax/peds_vax.parquet') %>%
  rename(birth_year = `Birth Year/Birth Cohort`, dim1=`Dimension Type`, urban=Dimension,vax_uptake=`Estimate (%)`, samp_size_vax=`Sample Size`) %>%
  collect() %>%
  filter(birth_year=='2016-2019' & dim1=='Urbanicity') %>%
  dplyr::select(Vaccine,Geography, Dose, dim1, vax_uptake,samp_size_vax, urban) %>%
  filter( Geography %in% state.name &
            (Vaccine %in% c('≥1 Dose MMR','≥1 Dose Varicella' ) | 
               (Vaccine=='DTaP' & Dose =='≥4 Doses') | 
               (Vaccine=='Hep A' & Dose =='≥2 Doses') | 
               (Vaccine=='Hep B' & Dose =='≥3 Doses') | 
               (Vaccine=='Hib' & Dose =='Full Series') | 
               (Vaccine=='PCV' & Dose =='≥4 Doses') 
            ) ) %>%
  mutate(urban= factor(urban, levels= c("Living In a Non-MSA", "Living In a MSA Non-Principal City","Living In a MSA Principal City"),labels=c('Rural','Smaller City', 'Larger City') ))

vax_insurance <- open_dataset('./Data/vax/peds_vax.parquet') %>%
  rename(birth_year = `Birth Year/Birth Cohort`, dim1=`Dimension Type`, insurance=Dimension,vax_uptake=`Estimate (%)`, samp_size_vax=`Sample Size`) %>%
  collect() %>%
  filter(birth_year=='2016-2019' & dim1=='Insurance Coverage') %>%
  dplyr::select(Vaccine,Geography, Dose, dim1, vax_uptake,samp_size_vax, insurance) %>%
  filter( Geography %in% state.name &
            (Vaccine %in% c('≥1 Dose MMR','≥1 Dose Varicella' ) | 
               (Vaccine=='DTaP' & Dose =='≥4 Doses') | 
               (Vaccine=='Hep A' & Dose =='≥2 Doses') | 
               (Vaccine=='Hep B' & Dose =='≥3 Doses') | 
               (Vaccine=='Hib' & Dose =='Full Series') | 
               (Vaccine=='PCV' & Dose =='≥4 Doses') 
            ) ) %>%
  mutate(insurance= factor(insurance, levels= c("Uninsured", "Any Medicaid","Private Insurance Only","Other"),labels=c('Uninsured','Medicaid', 'Private','Other') ))



```

```{r opioid_data}
#| context: data
#| include: false
#| cache: true
#| cache.extra: !expr file.info("data.csv")$mtime

# Interactive
df1 <- read_csv('./Data/Harmonized Opioid Overdose Datasets_01.23.2025.csv', show_col_types = FALSE) |>
  clean_names() |> 
  dplyr::filter(!(count %in% c(7777, 8888, 9999))) |> 
  dplyr::filter(!(crude_rate %in% c(7777, 8888, 9999))) |> 
  dplyr::filter(!(age_adjusted_rate %in% c(7777, 8888, 9999))) %>%
  mutate(qtr=as.numeric((gsub("Q",'', quarter))), 
         qtr =if_else(is.na(qtr),1,qtr),
         qdate= year+ qtr/4 -1/4
  )

opioid_path <- "./Data/Harmonized Opioid Overdose Datasets_01.23.2025.csv"

df_opioid <- read_csv(opioid_path) |>
  clean_names() |> 
  filter(!(count %in% c(7777, 8888, 9999))) |> 
  filter(!(crude_rate %in% c(7777, 8888, 9999))) |> 
  filter(!(age_adjusted_rate %in% c(7777, 8888, 9999))) |> 
  # Filter out rows without Quarter
  filter(!is.na(quarter)) |> 
  # Create year_quarter column
  mutate(year_quarter = paste(year, quarter))

#time series dataset
opioid_ts_ds  <- df_opioid %>%
  filter(drug=='All Opioids' & 
           characteristic=='Age' ) |> 
  group_by(dataset, state, year_quarter, level) %>%
  summarize(count=sum(count, na.rm = TRUE)) |>
  ungroup() |> 
  pivot_wider(names_from = level, values_from = count) |> 
  mutate(Overall = `25-44 Years` + `45-64 Years` + `65+ Years` + `<24 Years`) |> 
  pivot_longer(
    cols = where(is.numeric),
    names_to = "level",
    values_to = "count"
  ) |> 
  mutate(level = factor(level),
         level = fct_relevel(level, c("Overall", "65+ Years",
                                      "45-64 Years", "25-44 Years",
                                      "<24 Years"))) |> 
  filter(count > 0) 


# Static
opioid_od <- read_csv(opioid_path) %>%
  as.data.frame()


# Waffle Plots
subset <- read_csv("Data/Harmonized Opioid Overdose Datasets_02.05.2025.csv") |> 
  # Filter the placeholder numerical values.
  #filter(`Age Adjusted Rate` %!in% 7777 & `Age Adjusted Rate` %!in% 8888 & `Age Adjusted Rate` %!in% 9999) %>%
  filter(Count %!in% 7777 & Count %!in% 8888 & Count %!in% 9999) %>%
  
  # Removing the extraneous Drug and Year classes that are not common to the
  # SUDORS and CDC WONDER dataset.
  filter(Drug %!in% c("Naloxone"), Year %in% c(2020, 2021, 2022)) %>%
  
  # Filter the metadata settings.
  filter(State %in% "US", Quarter %in% NA, Setting %in% "All",
         `Underlying Cause of Death` %in% "Unintentional",
         Characteristic %in% "Not Stratified", Level %in% "N/A") %>%
  
  # Some random duplicates were detected. Not sure where these might have
  # been introduced or if the dataset came this way. This script is meant to
  # quickly deal with these.
  group_by(Dataset, Year, Drug) %>%
  mutate(Count = sum(Count)) %>%
  ungroup() %>%
  
  select(-`Crude Rate`, -`Age Adjusted Rate`, -Population) %>%
  distinct() %>%
  
  # Calculate the relative proportion of overdose events for a given Dataset
  # and Year.
  group_by(Dataset, Year) %>%
  mutate(Percentage = floor(Count/sum(Count)*100)) %>%
  ungroup() %>%
  
  # Waffle plots do not accept NA's or 0's. Remove them.
  mutate_at(c("Percentage"), ~replace_na(., 0)) %>%
  filter(!Percentage == 0) %>%
  
  # Organize the final table.
  .[with(., order(Dataset, Year, Percentage)), ] %>% `rownames<-`(NULL) %>%
  as.data.frame()


coloring <- data.frame("Drug"  = sort(unique(subset$Drug)),
                       "Color" = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99', '#e31a1c', '#fdbf6f','#ff7f00','#cab2d6','#6a3d9a','#ffff99')[1:length(sort(unique(subset$Drug)))]
                       #"Color" = c('#8dd3c7','#bebada','#fb8072','#80b1d3','#fdb462','#b3de69','#fccde5','#d9d9d9','#bc80bd','#ccebc5','#ffffb3')[1:length(sort(unique(subset$Drug)))]
                       )


waffle_2020 <- subset %>%
  filter(Year %in% 2020)

waffle_2021 <- subset %>%
  filter(Year %in% 2021)

waffle_2022 <- subset %>%
  filter(Year %in% 2022)
```

```{r}


#library(leaflet)
#https://r-graph-gallery.com/183-choropleth-map-with-leaflet.html
```

## Welcome to PopHIVE 
The Population Health Information and Visualization Exchange, hosted by the Yale School of Public Health. The goal of this initiative is to collect and present timely and informative data on population health from diverse sources, including traditional public health surveillance data, electronic health records, sewage testing, internet search trends, population surveys, and modeled outputs. We are just getting started and will be adding many more datasets and disease endpoints in the coming months.

------------------------------------------------------------------------

```{r rsv_plots}
#| context: server

########################################
##RSV
########################################
output$distPlotRSV3 <- renderPlotly({
  
  point_plot <- rsv_ts %>% filter(hhs_abbr==input$hhs.region.select &  epiyr == max(epiyr, na.rm=T) ) %>%
    filter(epiwk==max(epiwk))
  
  rsv_ts1 <- rsv_ts %>%
    filter(x==input$hhs.region.select) %>%
    ggplot(aes(x=epiwk, y=scaled_cases, group=as.factor(epiyr), color=as.factor(epiyr)))+
    geom_line()+
    theme_minimal()+
    xlab('Weeks since July')+
    ylab('RSV positive tests')+
    scale_colour_viridis_d() +
    geom_point(data=point_plot,aes(x=epiwk, y=scaled_cases) ,col='red') +
    guides(color=guide_legend(title="Season starting:"))
  
  
  ggplotly(rsv_ts1)
})


############################
#https://www.cdc.gov/wcms/vizdata/NCEZID_DIDRI/RSVStateLevelDownloadCSV.csv

output$distPlotRSV_dwh <- renderPlotly({
  
  # point_plot <- dwh %>% 
  #   dplyr::select(state,date, percent_visits_rsv,rsv_ww,hosp_rate) %>%
  #  filter(state==input$state.select2 ) %>%
  #  reshape2::melt(id.vars=c('state','date')) %>%
  #   group_by(variable) %>%
  #   filter(date==max(date))
  
  rsv_ts_comp <- dwh %>%
    filter(state==input$state.select2 & date >=as.Date('2023-07-01')) %>%
    ggplot()+
    geom_line(aes(x=date, y=percent_visits_rsv/max(percent_visits_rsv, na.rm=T), color='ED')) +
    geom_line(aes(x=date, y=rsv_ww/max(rsv_ww, na.rm=T), color='WasteW')) +
    geom_line(aes(x=date, y=hosp_rate/max(hosp_rate, na.rm=T),color='Hosp'), lty=2) +  
    theme_minimal()+
    xlab('Date')+
    ylab('RSV activity')+
    scale_color_manual(name = "RSV", values = c("ED" = "#e41a1c", "WasteW"="#377eb8", 'Hosp'='#4daf4a'  ))+
    ggtitle(input$state.select2)
  
  scale_colour_viridis_d() 
  
  
  ggplotly(rsv_ts_comp)
})


#hospitalization by age
output$distPlotRSV_hosp_age <- renderPlotly({
  
  rsv_ts_comp <- h1.age %>%
    filter(state==input$state.select3 & date >=as.Date('2023-07-01')) %>%
    
    ggplot()+
    geom_line(aes(x=date, y=scale_age, group_by=Level, color=Level)) +
    theme_minimal()+
    xlab('Date')+
    ylab('RSV activity')+
    scale_colour_viridis_d() +
    ggtitle(input$state.select3)
  
  ggplotly(rsv_ts_comp)
})

```

```{r pneumococcus_plots}
#| context: server 
#Pneumococcal disease
output$stPlot <- renderPlotly({
  p1 <- ipd1  %>%
    filter(st %in% c(input$select.st)) %>%
    ggplot(aes(x=year, y=N_IPD))+
    geom_line()+
    facet_wrap(~agec2, scales ='free', nrow=1) +
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, vjust = 1.0, hjust=1)) +
    geom_vline(xintercept=c(1999.5, 2009.5), lty=2, color='gray')+
    ylim(0,NA) +
    ggtitle(paste0('Trends in IPD caused by serotype ', input$select.st, ' in the US'))
  ggplotly(p1)
})


#serotype geography
output$stGeographyPlot <- renderPlotly({
  
  b2019 %>%
    filter(sero==input$select.st2 ) %>%
    ggplot( aes(fill=State, y=pct, x=sero)) +
    geom_bar(position="dodge", stat="identity")+
    scale_fill_viridis_d()+
    theme_minimal()+
    ylab('Proportion of IPD cases in the state') +
    ggtitle(input$select.st2)
})
```

```{r vax_plots}
#| context: server

#vaccine urbanicity
output$vax_urban_plot <- renderPlotly({
  
  plot_vax <- vax_urban %>% 
    filter(Vaccine==input$select.vax2) %>% mutate(vax_order=as.numeric(as.factor(Vaccine)), Vaccine_dose=as.factor(paste(Vaccine,Dose)) ) %>% 
    ggplot(aes(x = vax_uptake, y = urban, 
               text=paste0(Geography, "<br>", 
                           Vaccine_dose, "<br>",
                           vax_uptake, "%" ))) +
    geom_jitter(aes(size = samp_size_vax, color = vax_uptake), height = 0.1, width = 0, alpha = 0.7) + 
    scale_size(range = c(2, 5), name="Uptake (%)")+ scale_color_viridis_c(option = "viridis") +
    labs(title = "Vaccination Rates by State, 35 months of age", x = "Vaccination Rate (%)", y = '') + 
    theme_minimal() + 
    scale_y_discrete(expand = expansion(mult = c(.1, .1))) + # Add more space between rows 
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5), axis.title.y = element_text(angle = 0, vjust = 0.5))
  
  ggplotly(plot_vax, tooltip='text') #display the plot
  
  
})

#vaccine, by insurance state
output$vax_insurance_plot <- renderPlotly({
  
  plot_ins <- vax_insurance %>% 
    filter(Vaccine==input$select.vax2) %>% mutate(vax_order=as.numeric(as.factor(Vaccine)),  
                                                  Vaccine_dose=as.factor(paste(Vaccine,Dose)) ) %>% 
    ggplot(aes(x = vax_uptake, y = insurance, 
               text=paste0(Geography, "<br>", 
                           Vaccine_dose, "<br>",
                           vax_uptake, "%" ))) +
    geom_jitter(aes(size = samp_size_vax, color = vax_uptake), height = 0.1, width = 0, alpha = 0.7) + 
    scale_size(range = c(2, 5), name="Uptake (%)")+ scale_color_viridis_c(option = "viridis") +
    labs(title = "Vaccination Rates by State, 35 months of age", x = "Vaccination Rate (%)", y = '') + 
    theme_minimal() + 
    scale_y_discrete(expand = expansion(mult = c(.1, .1))) + # Add more space between rows 
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5), axis.title.y = element_text(angle = 0, 
                                                                                                        vjust = 0.5))
  
  ggplotly(plot_ins, tooltip='text') #display the plot
  
})
```

```{r opioid_plots}
#| context: server


output$distPlot1 <- renderPlotly({
  
  p1 <- df1 %>%
    dplyr::filter(drug=='All Opioids' & characteristic=='Age' & dataset=='AHRQ' & state == input$state.select ) %>%
    group_by(qdate, level) %>%
    summarize(count=sum(count)) %>% #combine outpatient nad inpatient
    ggplot(aes(x=qdate, y=count, group=level, color=level)) +
    theme_minimal() +
    geom_line()
  ggplotly(p1)
})

output$distPlot2 <- renderPlotly({
  
  p1 <- df1 %>%
    dplyr::filter(drug=='All Opioids' & characteristic=='Age' & dataset=='NCHS' & state == input$state.select ) %>%
    group_by(qdate, level) %>%
    summarize(count=sum(count)) %>% #combine outpatient nad inpatient
    ggplot(aes(x=qdate, y=count, group=level, color=level)) +
    theme_minimal() +
    geom_line()
  ggplotly(p1)
})

## Opioid
df_time_series_interactive <- reactive({
  df_opioid %>%
    filter(drug=='All Opioids' & 
             characteristic=='Age' &
             state == input$state) |> 
    group_by(dataset, state, year_quarter, level) %>%
    summarize(count=sum(count, na.rm = TRUE)) |>
    ungroup() |> 
    pivot_wider(names_from = level, values_from = count) |> 
    mutate(Overall = `25-44 Years` + `45-64 Years` + `65+ Years` + `<24 Years`) |> 
    pivot_longer(
      cols = where(is.numeric),
      names_to = "level",
      values_to = "count"
    ) |> 
    mutate(level = factor(level),
           level = fct_relevel(level, c("Overall", "65+ Years",
                                        "45-64 Years", "25-44 Years",
                                        "<24 Years"))) |> 
    filter(count > 0) 
})

output$opioid_time_series_interactive_1 <- renderPlotly({ 
  # User must provide inputs
  req(input$state)
  # Dataset must have rows
  req(nrow(df_time_series_interactive()) > 0)
  
  p1 <- df_time_series_interactive() %>%
    ggplot(aes(x = year_quarter, y = count, 
               group = level, color = level,
               text = paste0("Quarter: ", year_quarter, "\n",
                             "Count: ", scales::comma(count)))) +
    geom_line() +
    labs(x = NULL,
         y = "Count",
         color = "Age Category") +
    scale_x_discrete(breaks = c("2016 Q1", "2017 Q1", "2018 Q1",
                                "2019 Q1", "2020 Q1", "2021 Q1", 
                                "2022 Q1")) +
    scale_y_continuous(labels = scales::comma) +
    scale_color_manual(values = yale_colors) + # Apply Yale branding colors
    facet_wrap(~dataset) +
    theme_minimal(base_size = 15) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggplotly(p1, tooltip = c("text"))
}) 



output$opioid_time_series_interactive_2 <- renderPlotly({
  
  p <- df_opioid %>%
    filter(drug=='All Opioids' & 
             characteristic=='Sex' &
             state == "US") %>%
    group_by(dataset, state, year_quarter, level) %>%
    summarize(count=sum(count, na.rm = TRUE)) %>%
    ungroup() %>%
    filter(count > 0) %>%
    ggplot(aes(x = year_quarter, y = count, 
               group = level, color = level,
               text = paste0("Quarter: ", year_quarter, "\n",
                             "Count: ", scales::comma(count)))) +
    geom_line() +
    labs(x = NULL,
         y = "Count",
         color = NULL,
         title = "National Opioid Overdose Count") +
    scale_x_discrete(breaks = c("2016 Q1", "2017 Q1", "2018 Q1",
                                "2019 Q1", "2020 Q1", "2021 Q1", 
                                "2022 Q1")) +
    scale_y_continuous(labels = scales::comma) +
    scale_color_manual(values = yale_colors) + # Apply Yale branding colors
    facet_wrap(~dataset) +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45,  hjust = 1))
  
  ggplotly(p, tooltip = "text")
})

output$opioid_bar_graph_2020 <- renderPlotly({
  
  p <- opioid_od %>%
    # Filter the placeholder numerical values.
    filter(`Age Adjusted Rate` %!in% 7777 & `Age Adjusted Rate` %!in% 8888 & `Age Adjusted Rate` %!in% 9999) %>%
    
    # Change the date named in the plot title.
    filter(Year %in% 2020) %>%
    
    # Filter the metadata settings.
    filter(State %in% "US", Quarter %in% NA, Setting %in% "All",
           `Underlying Cause of Death` %in% "Unintentional",
           Characteristic %in% "Not Stratified", Level %in% "N/A") %>%
    mutate(Drug = if_else(Drug == "Mental and behavioural disorders due to use of opioids, acute intoxication",
                          "Mental and behavioural disorders",
                          Drug)) %>%
    
    # Plot settings and features.
    ggplot(data = ., aes(x = Drug, y = `Age Adjusted Rate`,
                         text = paste0("Drug: ", Drug, "<br>",
                                       "Age Adjusted Rate: ", `Age Adjusted Rate`))) +
    geom_bar(stat = "identity", position = "dodge", aes(fill = Dataset, color = Dataset)) +
    labs(title = "National Opioid Overdose Rate\nby Types of Opioid and Polysubstance in 2020",
         subtitle = "Underlying Cause of Death: Unintentional Setting: All.",
         x = "", y = "Age-Adjusted Rate (per 100,000)") +
    scale_color_manual(values = yale_colors) + # Apply Yale branding colors
    scale_fill_manual(values = yale_colors) + # Apply Yale branding colors
    theme_minimal(base_size = 13) + 
    theme(axis.text.x = element_text(angle = 90,  hjust = 1))
  
  ggplotly(p, tooltip = "text")
})

output$opioid_bar_graph_2021 <- renderPlotly({
  
  
  p <- opioid_od %>%
    # Filter the placeholder numerical values.
    filter(`Age Adjusted Rate` %!in% 7777 & `Age Adjusted Rate` %!in% 8888 & `Age Adjusted Rate` %!in% 9999) %>%
    
    # Change the date named in the plot title.
    filter(Year %in% 2021) %>%
    
    # Filter the metadata settings.
    filter(State %in% "US", Quarter %in% NA, Setting %in% "All",
           `Underlying Cause of Death` %in% "Unintentional",
           Characteristic %in% "Not Stratified", Level %in% "N/A") %>%
    mutate(Drug = if_else(Drug == "Mental and behavioural disorders due to use of opioids, acute intoxication",
                          "Mental and behavioural disorders",
                          Drug)) %>%
    
    # Plot settings and features.
    ggplot(data = ., aes(x = Drug, y = `Age Adjusted Rate`,
                         text = paste0("Drug: ", Drug, "<br>",
                                       "Age Adjusted Rate: ", `Age Adjusted Rate`))) +
    geom_bar(stat = "identity", position = "dodge", aes(fill = Dataset, color = Dataset)) +
    labs(title = "National Opioid Overdose Rate by\nTypes of Opioid and Polysubstance in 2021",
         subtitle = "Underlying Cause of Death: Unintentional Setting: All.",
         x = "", y = "Age-Adjusted Rate\n(per 100,000)") +
    scale_color_manual(values = yale_colors) + # Apply Yale branding colors
    scale_fill_manual(values = yale_colors) + # Apply Yale branding colors
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 90,  hjust = 1))
  
  ggplotly(p, tooltip = "text")
})

output$opioid_bar_graph_2022 <- renderPlotly({
  
  
  p <- opioid_od %>%
    # Filter the placeholder numerical values.
    filter(`Age Adjusted Rate` %!in% 7777 & `Age Adjusted Rate` %!in% 8888 & `Age Adjusted Rate` %!in% 9999) %>%
    
    # Change the date named in the plot title.
    filter(Year %in% 2022) %>%
    
    # Filter the metadata settings.
    filter(State %in% "US", Quarter %in% NA, Setting %in% "All",
           `Underlying Cause of Death` %in% "Unintentional",
           Characteristic %in% "Not Stratified", Level %in% "N/A") %>%
    mutate(Drug = if_else(Drug == "Mental and behavioural disorders due to use of opioids, acute intoxication",
                          "Mental and behavioural disorders",
                          Drug)) %>%
    
    # Plot settings and features.
    ggplot(data = ., aes(x = Drug, y = `Age Adjusted Rate`,
                         text = paste0("Drug: ", Drug, "<br>",
                                       "Age Adjusted Rate: ", `Age Adjusted Rate`))) +
    geom_bar(stat = "identity", position = "dodge", aes(fill = Dataset, color = Dataset)) +
    labs(title = "National Opioid Overdose Rate\nby Types of Opioid and Polysubstance in 2022",
         subtitle = "Underlying Cause of Death: Unintentional Setting: All.",
         x = "", y = "Age-Adjusted Rate\n(per 100,000)") +
    scale_color_manual(values = yale_colors) + # Apply Yale branding colors
    scale_fill_manual(values = yale_colors) + # Apply Yale branding colors
    theme_minimal(base_size = 13) + 
    theme(axis.text.x = element_text(angle = 90,  hjust = 1))
  
  ggplotly(p, tooltip = "text")
  
})



output$ahrq_map <- renderPlotly({
  
  
  # -----------------------------
  # US Map plot.
  
  # Code to find mathes between AHRQ and CDC WONDER
  #opioid_od[opioid_od$Dataset %in% "AHRQ", "Setting"] %>% unique()
  
  # Generate a side-by-side set of plots showing "Drug = All Opioids" counts
  # by state in 2022. Toggle the setting to show either inpatient or ER.
  
  
  # Counts in AHRQ dataset.
  ahrq_map_plot <- opioid_od %>%
    # Filter the placeholder numerical values.
    filter(Count %!in% 7777 & Count %!in% 8888 & Count %!in% 9999) %>%
    
    # Switch between the two settings, and change the subtitle name.
    filter(Setting %in% "Medical Facility - Inpatient") %>%
    #filter(Setting %in% "Medical Facility - Outpatient or ER") %>%
    
    # Filter the metadata settings.
    filter(Dataset %in% "AHRQ", State %in% datasets::state.name, 
           Year %in% 2022, Quarter %in% NA,
           `Underlying Cause of Death` %in% "All", Drug %in% "All Opioids",
           Characteristic %in% "Not Stratified", Level %in% "N/A") %>%
    
    # plot_usmap() requires specific nomenclature for the column with states.
    rename(state = State) %>%
    
    # Plot settings and features.
    plot_usmap(data = ., values = "Count", color = "#00356B",) + 
    scale_fill_gradientn(
      colors = yale_gradient,   # Use Yale branding colors in gradient
      name = "Count (2022)", 
      labels = scales::comma    # Format labels with commas
    ) + 
    labs(title = "AHRQ") +
    theme(legend.position = "",
          plot.title = element_text(size = 14),  # Increase title size
          axis.ticks = element_blank(),
          axis.text = element_blank(),
          legend.text = element_text(size = 12),               # Adjust legend text
          legend.title = element_text(size = 14))
  
  # Add a custom tooltip with comma formatting
  ahrq_map_plot <- ahrq_map_plot +
    aes(text = paste0("State: ", state, "<br>",
                      "Count: ", scales::comma(Count)))
  
  # Convert to plotly
  ggplotly(ahrq_map_plot, tooltip = "text")
})


output$wonder_map <- renderPlotly({
  # Counts in CDC WONDER dataset.
  wonder_map_plot <- opioid_od %>%
    # Filter the placeholder numerical values.
    filter(Count %!in% 7777 & Count %!in% 8888 & Count %!in% 9999) %>%
    
    # Switch between the two settings, and change the subtitle name.
    filter(Setting %in% "Medical Facility - Inpatient") %>%
    #filter(Setting %in% "Medical Facility - Outpatient or ER") %>%
    
    # Filter the metadata settings.
    filter(Dataset %in% "CDC WONDER", State %in% datasets::state.name, 
           Year %in% 2022, Quarter %in% NA,
           `Underlying Cause of Death` %in% "All", Drug %in% "All Opioids",
           Characteristic %in% "Not Stratified", Level %in% "N/A") %>%
    
    # plot_usmap() requires specific nomenclature for the column with states.
    rename(state = State) %>%
    
    # Plot settings and features.
    plot_usmap(data = ., values = "Count", color = "#00356B",) + 
    scale_fill_gradientn(
      colors = yale_gradient,   # Use Yale branding colors in gradient
      name = "Count (2022)", 
      labels = scales::comma    # Format labels with commas
    ) + 
    labs(title = "\ \ \ \ \ \ CDC WONDER") +
    theme(legend.position = "right",
          plot.title = element_text(size = 14),  # Increase title size
          axis.ticks = element_blank(),
          axis.text = element_blank(),
          legend.text = element_text(size = 14),               # Adjust legend text
          legend.title = element_text(size = 16))
  
  # Add a custom tooltip with comma formatting
  wonder_map_plot <- wonder_map_plot +
    aes(text = paste0("State: ", state, "<br>",
                      "Count: ", scales::comma(Count)))
  
  ggplotly(wonder_map_plot, tooltip = "text")
  
})

output$waffle_plot <- renderPlot({
  
  
  # Generate the waffle plots.
  w1 <- waffle_2020 %>%
    ggplot(aes(fill = Drug, values = Percentage)) +
    geom_waffle(n_rows = 5, size = 0.33, colour = "white") +
    facet_wrap(~Dataset+Year) +
    # Add the color scheme specific for the drugs present in the SUDORS subset.
    scale_fill_manual(name = NULL, 
                      labels = coloring[coloring$Drug %in% unique(waffle_2020$Drug), "Drug"],
                      values = coloring[coloring$Drug %in% unique(waffle_2020$Drug), "Color"]) +
    coord_equal() + theme_void(base_size = 16) + 
    theme(legend.position = "none")
  
  
  w2 <- waffle_2021 %>%
    ggplot(aes(fill = Drug, values = Percentage)) +
    geom_waffle(n_rows = 5, size = 0.33, colour = "white") +
    facet_wrap(~Dataset+Year) +
    # Add the color scheme specific for the drugs present in the CDC WONDER subset.
    scale_fill_manual(name = NULL, 
                      labels = coloring[coloring$Drug %in% unique(waffle_2021$Drug), "Drug"],
                      values = coloring[coloring$Drug %in% unique(waffle_2021$Drug), "Color"]) +
    coord_equal() + theme_void(base_size = 16) + 
    theme(legend.position = "none")
  
  
  w3 <- waffle_2022 %>%
    ggplot(aes(fill = Drug, values = Percentage)) +
    geom_waffle(n_rows = 5, size = 0.33, colour = "white") +
    facet_wrap(~Dataset+Year) +
    # Add the color scheme specific for the drugs present in the CDC WONDER subset.
    scale_fill_manual(name = NULL, 
                      labels = coloring[coloring$Drug %in% unique(waffle_2022$Drug), "Drug"],
                      values = coloring[coloring$Drug %in% unique(waffle_2022$Drug), "Color"]) +
    coord_equal() + theme_void(base_size = 16) + 
    theme(legend.position = "bottom", legend.title = element_blank())
  
  
  iron(w1, w2, w3)
})



```

## Respiratory syncytial virus (RSV) in the United States

RSV, a virus that infects the lungs, is an important cause of hospitalizations in infants and older adults. RSV rates tend to start increasing each year in the southeastern US in early autumn, with waves occurring later to the north and west over several months.

::: panel-tabset
### Composite trends

```{r}
#| panel: toolbar

selectInput("state.select2", "State:", 
            choices=state.name, selected='New York')
```

```{r}
#| panel: fill

plotlyOutput("distPlotRSV_dwh")

```

Viral levels in the community can be measured in different ways. This plot shows how relying on data from the emergency department (ED), hospitalizations, and sewage surveillance can yield different estimates of epidemic timing. This could be because of who contributes to the different datasets: different age groups could contribute different aount of information to each line, and the location of the EDs and the location of the sewage sampling

### Trends in positive tests

```{r}
#| panel: toolbar

selectInput("hhs.region.select", "HHS Region:", 
            choices=unique(rsv_ts$hhs_abbr), selected='CT,ME,MA,NH,RI,VT')
```

```{r}
#| panel: fill
plotlyOutput("distPlotRSV3")

```

These data come from the NREVSS surveillance system, which is comprised of laboratories around the US

### Age patterns in hospitalization

```{r}
#| panel: toolbar

selectInput("state.select3", "State:", 
            choices=unique(h1.age$state), selected='New York')
```

```{r}
#| panel: fill

plotlyOutput("distPlotRSV_hosp_age")

```

The hospitalization data comes from the RSV-NET, a CDC-supported network of sites around the US that tracks hospitalizations from RSV, influenza, and COVID-19.

### Current national map (ED)

```{r}
#| panel: fill

i=length(dates)-1
d1 <- open_dataset('./Data/NSSP_detailed.parquet') %>%
  filter(county!='All' & week_end==dates[i] ) %>%
  mutate(
    percent_visits_rsv=if_else(percent_visits_rsv>1,1,percent_visits_rsv) ,
    
    
    percent_visits_combined=if_else(percent_visits_combined>10,10,percent_visits_combined) ) %>%
  collect()

p1 <- usmap::plot_usmap(regions='counties', data=d1,values='percent_visits_rsv' )   +
  labs(title = paste0('RSV ED visits ', dates[i])) + 
  theme(panel.background = element_rect(color = "white", fill = "white")) +
  scale_fill_stepsn('RSV', breaks= seq(0,1, by=0.1), limits = c(0,1),
                    colors=viridis_pal()(10),
                    guide = guide_colorsteps(even.steps = FALSE))

print(p1)
```

The maps shows the proportion of ED visits in each jurisdiction caused by RSV. Some states have county-level data available, while others have state-level data only. This includes data aggregated across all age groups. Ideally this would be broken out separately for infants and older adults. The data come from the national syndromic surveillance network.

### Regional trends (NREVSS)

```{r}
heat1 <- rsv1_tests %>%
  mutate(date= as.Date(substr(mmwrweek_end,1,10 ), '%m/%d/%Y'),
         postdate= as.Date(substr(posted,1,10 ), '%m/%d/%Y')) %>%
  filter(postdate==max(postdate)) %>% 
  ungroup() %>%
  filter(level != 'National') %>%
  group_by(level ) %>%
  left_join(key, by=c('level'='Group.1')) %>%
  mutate(scaled_cases = pcr_detections/max(pcr_detections),
         hhs_abbr=x)

# by each district
p2 <- ggplot(heat1, aes(date, x, fill = scaled_cases))+
  geom_tile()+ 
  scale_fill_distiller(name = "PCR Positives", 
                       palette = "Blues", 
                       direction = 1,
                       breaks = pretty_breaks())+
  theme(axis.text.y = element_text(size = 3))+
  labs(x = "Date", y = "HHS Region", title ="RSV detections NREVSS")+
  guides(fill="none")+
  theme_minimal()
p2
```

This dataset has viral test results from a network of laboratories around the US (NREVSS). The location of the labs doesn't necesarilycover the whole population in a state or region. The CDC only publishes these data at the region level

### Trends in Google searches

```{r}
#| panel: fill

print('Coming soon!')
```

People searching for information on RSV often correlates well with actual clinical activity for RSV.
:::

------------------------------------------------------------------------

## Pneumococcal disease in the United States

Pneumococcus is an important bacterial pathogen that causes severe disease like meningitis, blood-stream infections, and pneumonia. It also is a major cause of ear infections in children. Vaccines (PCVs) have been in use in the USA since 2000, with a switch to a new PCV that covered more strains in 2010. The introduction of these vaccines has largely led to the elimination of vaccine-targeted serotypes and smaller increases in disease caused by other serotypes. Some vaccine-targeted serotypes have recently re-emerged (e.g., 19F) for reasons that are still not clear.

::: panel-tabset
### Trends in serotype frequency

```{r}
#| panel: toolbar
selectInput("select.st",
            "Select serotype:",
            all.sts,
            selected='19F')
```

```{r}
#| panel: fill
plotlyOutput("stPlot")

```

This shows trends in Invasive Pneumococcal Disease caused by different serotypes, as reported to the Active Bacterial Core Surevillance system (ABCs) at CDC. This is a network of 10 sites around the US.

### Change in Invasive pneumococcal Disease 1998-2022

```{r}
#| include: false

pcv7 <- c('4','6B','9V','14','18C','19F','23F')

pcv10gsk <- c(pcv7,'1','5','7F')

pcv13 <- c(pcv7, '1','3','5','6A','7F','19A')

s1 <- readRDS('./Data/pneumococcus/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
  group_by(st) %>%
  mutate(cum_N= sum(N_IPD)) %>%
  filter(cum_N>100) %>%
  ungroup() %>%
  mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT'))
s2_pre<- s1 %>%
  filter(year %in% c(1998, 1999)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
  group_by(st) %>%
  summarize(N_IPD_pre=mean(N_IPD)) %>%
  ungroup() %>%
  tidyr::complete(st,  fill=list(N_IPD_pre=0))  #fills 0

s2_pre13<- s1 %>%
  filter(year %in% c(2008, 2009)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
  group_by(st) %>%
  summarize(N_IPD_pre=mean(N_IPD)) %>%
  ungroup() %>%
  tidyr::complete(st,  fill=list(N_IPD_pre=0))  #fills 0

s2<- s1 %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>% #sum across age group
  ungroup() %>%
  tidyr::complete(st, year, fill=list(N_IPD=0))%>%  #fills 0
  left_join(s2_pre, by='st') %>%
  mutate(N_IPD_pre = if_else(is.na(N_IPD_pre),0, N_IPD_pre) ,
         logRR = log((N_IPD+1)/(N_IPD_pre+1) )) 

max_RR <- s2 %>%
  group_by(st) %>%
  summarize(max_RR = max(logRR)) %>%
  arrange((max_RR))

s2$st <- factor(s2$st, levels = max_RR$st)

df_wide <- s2 %>%
  dplyr::select(year, st, logRR) %>%
  tidyr::pivot_wider(names_from = year, values_from = logRR)


df_vt <- df_wide %>%
  mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT')) %>%
  filter(pcv13st =='VT') %>%
  dplyr::select(-pcv13st) 

df_nvt <- df_wide %>%
  mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT')) %>%
  filter(pcv13st =='NVT') %>%
  dplyr::select(-pcv13st) 

rr_nvt <- s2 %>%
  mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT')) %>%
  filter(pcv13st =='NVT') %>%
  ggplot( aes(x = factor(year), y = st, fill = logRR)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(x = "Year", y =
         "serotype", fill = "log RR") +
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, hjust=1))

rr_vt <- s2 %>%
  mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT')) %>%
  filter(pcv13st =='VT') %>%
  ggplot( aes(x = factor(year), y = st, fill = logRR)) +
  geom_tile() +
  #     guides(fill="none")+
  
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(x = "Year", y =
         "serotype", fill = "log RR") +
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r}
#| panel: fill
subplot(ggplotly(rr_vt), ggplotly(rr_nvt),  margin = 0.04)
```

This shows the decreases in serotypes targeted by PCV13 (left) and increases in the occurrence of non-vaccine serotypes (right) since 1998/1999

### Geographic variation in serotype frequency

Some serotypes have a similar distribution across the US, while others (e.g., serotype 4) have distinct geographic patterns. serotype 4 has been largely confined to the western US in recent years and is largely found among people who are homeless. Data from 2019 IPD (Beall et al, JID 2022 10.1093/infdis/jiac058)

```{r}
#| panel: toolbar
selectInput("select.st2",
            "Select serotype:",
            unique(b2019$sero),
            selected='19F')
```

```{r}
#| panel: fill
plotlyOutput("stGeographyPlot")

```

### Comparison of invasive pneumococcal disease and pneumonia

```{r}
ipd1 <- readRDS('./Data/pneumococcus/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st),
          st = if_else(st %in% c('15B','15C'), '15BC',st),
          if_else(st %in% c('6A','6C'), '6AC',st)
  ) %>%
  filter(year %in% c(2019,2020) & agec %in% c('Age 50-64','Age 65+')) %>%
  group_by(st) %>%
  summarize( N_IPD= sum(N_IPD))

uad <- readxl::read_excel('./Data/pneumococcus/SSUAD/ramirez_ofid_2025_ofae727.xlsx') %>%
  mutate(N_SSUAD= over65 + a50_64_with_indication + a50_64_no_indication ) %>%
  full_join(ipd1, by='st') %>%
  filter(!is.na(N_SSUAD) & !is.na(N_IPD))

ggplot(uad, aes(x=N_IPD, y=N_SSUAD, label=st)) +
  geom_point()+
  geom_text( vjust = 1)+
  theme_classic() +
  ggtitle('IPD vs pneumonia 50+ years in US')
```

Most often, pneumococcal disease is diagnosed by isolating the bacteria from blood or cerebrospinal fluid. It is harder to determine what serotypes cause pneumonia. This study from Louisville, Kentucky found that the serotypes that are the most common causes of severe blood stream infections are also the most common causes of pneumonia. pneumonia [Ramirez et al OFID 2025](https://academic.oup.com/ofid/article/12/1/ofae727/7926903)
:::

------------------------------------------------------------------------

## Vaccination

Vaccination of children across the US, by state and vaccine. These data come from the [National Immunization Survey](https://data.cdc.gov/Child-Vaccinations/Vaccination-Coverage-among-Young-Children-0-35-Mon/fhky-rtsk/about_data) The size of the circle is proportional to the number of children participating in the survey. Caution should be used when interpreting these survey data.

::: panel-tabset
### Uptake by state

```{r, fig.width=5, fig.height=7}
plot_vax <- vax_age %>% mutate(vax_order=as.numeric(as.factor(Vaccine)), Vaccine_dose=as.factor(paste(Vaccine,Dose)) ) %>% 
  ggplot(aes(x = vax_uptake, y = Vaccine_dose, 
             text=paste0(Geography, "<br>", 
                         Vaccine_dose, "<br>",
                         'Age: ',age, "<br>", 
                         vax_uptake, "%" ))) +
  geom_jitter(aes(size = samp_size_vax, color = vax_uptake), height = 0.25, width = 0, alpha = 0.7) + 
  scale_size(range = c(2, 5), name="Uptake (%)")+ scale_color_viridis_c(option = "viridis") +
  labs(title = "Vaccination Rates by State, 35 months of age", x = "Vaccination Rate (%)", y = '') + 
  theme_minimal() + 
  scale_y_discrete(expand = expansion(mult = c(.1, .1))) + # Add more space between rows 
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5), axis.title.y = element_text(angle = 0, vjust = 0.5))

ggplotly(plot_vax, tooltip='text')
```

### Uptake by urbanicity

Variation in vaccine uptake based on how urban an area is.

```{r}
#| panel: toolbar

selectInput("select.vax2", "Vaccine:", 
            choices=unique(vax_urban$Vaccine), selected='PCV')
```

```{r}
#| panel: fill
#| layout-ncol: 1
#| height: 1200px

plotlyOutput("vax_urban_plot")

```

Variation in vaccine uptake based on how urban an area is. Uptake in non-urban, smaller urban, larger urban locations

### Uptake based on insurance status

```{r}
#| panel: fill
#| layout-ncol: 1
#| height: 1200px

plotlyOutput("vax_insurance_plot")

```

Variation in vaccine uptake rates based on insurance status
:::

------------------------------------------------------------------------

## Opioids in the United States

::: panel-tabset
### Trends in national healthcare incidents

```{r}
#| panel: fill
plotlyOutput("opioid_time_series_interactive_2")

```

### Map of opioid overdoses

AHRQ hospitalizations

```{r}
#| panel: fill
plotlyOutput("ahrq_map")

```

CDC WONDER

```{r}
#| panel: fill
plotlyOutput("wonder_map")

```

### Type of opioid

```{r}
#| panel: fill
plotOutput("waffle_plot")
```



:::

------------------------------------------------------------------------

## Documentation

### Notes on data sources

Coming soon!

### Acknowledgements

We thank all of the organizations and individuals who make their data available for health monitoring and research.

Contributors include Shelby Golden, Howard Baik, Maurice Dalton, Megan Ranney, Bhramar Mukherjee, Katelyn Jetelina, Ron Borzekowski, Anne Zink, and Dan Weinberger

Questions or suggestions? Contact daniel dot weinberger at yale dot edu
